---
title: "robust regressions with morphologic measures"
author: "Francesco Pisu"
date: "11/15/2021"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
---

# Imports, paths and data loading
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# wrangling packages
library(here) # here makes a project transportable
library(janitor) # clean_names
library(readxl) # read excel, duh!
library(data.table) # magical data frames
library(magrittr) # pipes
library(stringr) # string functions
library(forcats) # factor functions
library(dplyr)
library(openxlsx) # write xlsx
library(purrr)

# analysis packages
library(emmeans) # the workhorse for inference
library(nlme) # gls and some lmm
library(lme4) # linear mixed models
library(lmerTest) # linear mixed model inference
library(afex) # ANOVA linear models
library(glmmTMB) # generalized linear models
library(MASS) # negative binomial and some other functions
library(car) # model checking and ANOVA
library(DHARMa) # model checking
library(RVAideMemoire)
library(robustbase)


# graphing packages
library(ggsci) # color palettes
library(ggpubr) # publication quality plots
library(ggforce) # better jitter
library(ggeffects)
library(visreg)
library(cowplot) # combine plots
library(knitr) # kable tables
library(kableExtra) # kable_styling tables
library(ggiraphExtra)
library(patchwork)

# ggplot_the_model.R packages not loaded above
library(insight)
library(lazyWeave)


# use here from the here package
here <- here::here
# use clean_names from the janitor package
clean_names <- janitor::clean_names

# load functions used by this text written by me
# ggplot_the_model.R needs to be in the folder "R"
# if you didn't download this and add to your R folder in your
# project, then this line will cause an error
source_path <- here("R", "ggplot_the_model.R")
source(source_path)

data_folder <- "data"
image_folder <- "figures"
output_folder <- "results"

file_name <- "df_with_stats.xlsx"
file_path <- here(data_folder, file_name)
df <- read_excel(file_path) %>% data.table()
df$gender <- ifelse(df$sex == 'M', 0, 1)
```

# Utilities
```{r}
print_coefs <- function(model) {
  model_coeff <- summary(model) %>%
    coef()
  model_confint <- confint(model)
  model_coeff <- cbind(model_coeff, model_confint)
  model_coeff %>% kable(digits=c(4, 4, 2, 3, 4, 4))
}
```

# Correlations

## Density ~ Age
```{r}
# Retrieve structures' names
tracts <- unique(df$structure)

tract_names <- c()

# not adjusted
r <- c()
ci_low <- c()
ci_up <- c()
P <- c()

for (i in 1:length(tracts)) {
  tract_name <- tracts[i]
  print(tract_name)
  tract <- df[structure == tract_name]
  
  set.seed(1234)
  c1 <- RVAideMemoire::spearman.ci(tract$age, tract$vol_length_ratio)
  c2 <- cor.test(tract$age, tract$vol_length_ratio, method='spearman', exact=FALSE)
  
  tract_names <- c(tract_names, tract_name)

  # correlation statistics
  r <- c(r, round(c1$estimate, 6))
  ci_low <- c(ci_low, round(c1$conf.int['Inf'], 5))
  ci_up <- c(ci_up, round(c1$conf.int['Sup'], 5))
  P <- c(P, round(c2$p.value, 4))
}
```

```{r}
cor_coefs <- data.frame(tract_names, r, ci_low, ci_up, P)
write.xlsx(cor_coefs, 'spearman_coefs_age_density.xlsx')
```

## Volume ~ Age
```{r}
# Retrieve structures' names
tracts <- unique(df$structure)

tract_names <- c()

# not adjusted
r <- c()
ci_low <- c()
ci_up <- c()
P <- c()

for (i in 1:length(tracts)) {
  tract_name <- tracts[i]
  print(tract_name)
  tract <- df[structure == tract_name]
  
  set.seed(1234)
  c1 <- RVAideMemoire::spearman.ci(tract$age, tract$volume)
  c2 <- cor.test(tract$age, tract$volume, method='spearman', exact=FALSE)
  
  tract_names <- c(tract_names, tract_name)
  
  # correlation statistics
  r <- c(r, round(c1$estimate, 6))
  ci_low <- c(ci_low, round(c1$conf.int['Inf'], 5))
  ci_up <- c(ci_up, round(c1$conf.int['Sup'], 5))
  P <- c(P, round(c2$p.value, 4))
}
```

```{r}
cor_coefs <- data.frame(tract_names, r, ci_low, ci_up, P)
write.xlsx(cor_coefs, 'spearman_coefs_age_volume.xlsx')
```

## Total Length ~ Age
```{r}
# Retrieve structures' names
tracts <- unique(df$structure)

tract_names <- c()

# not adjusted
r <- c()
ci_low <- c()
ci_up <- c()
P <- c()

for (i in 1:length(tracts)) {
  tract_name <- tracts[i]
  print(tract_name)
  tract <- df[structure == tract_name]
  
  set.seed(1234)
  c1 <- RVAideMemoire::spearman.ci(tract$age, tract$total_length)
  c2 <- cor.test(tract$age, tract$total_length, method='spearman', exact=FALSE)
  
  tract_names <- c(tract_names, tract_name)

  # correlation statistics
  r <- c(r, round(c1$estimate, 6))
  ci_low <- c(ci_low, round(c1$conf.int['Inf'], 5))
  ci_up <- c(ci_up, round(c1$conf.int['Sup'], 5))
  P <- c(P, round(c2$p.value, 4))
}
```

```{r}
cor_coefs <- data.frame(tract_names, r, ci_low, ci_up, P)
write.xlsx(cor_coefs, 'spearman_coefs_age_total_length.xlsx')
```

## Avg. Length ~ Age
```{r}
# Retrieve structures' names
tracts <- unique(df$structure)

tract_names <- c()

# not adjusted
r <- c()
ci_low <- c()
ci_up <- c()
P <- c()

for (i in 1:length(tracts)) {
  tract_name <- tracts[i]
  print(tract_name)
  tract <- df[structure == tract_name]
  
  set.seed(1234)
  c1 <- RVAideMemoire::spearman.ci(tract$age, tract$avg_streamline_length)
  c2 <- cor.test(tract$age, tract$avg_streamline_length, method='spearman', exact=FALSE)
  
  tract_names <- c(tract_names, tract_name)

  # correlation statistics
  r <- c(r, round(c1$estimate, 6))
  ci_low <- c(ci_low, round(c1$conf.int['Inf'], 5))
  ci_up <- c(ci_up, round(c1$conf.int['Sup'], 5))
  P <- c(P, round(c2$p.value, 4))
}
```

```{r}
cor_coefs <- data.frame(tract_names, r, ci_low, ci_up, P)
write.xlsx(cor_coefs, 'spearman_coefs_age_avg_length.xlsx')
```

# Regressions

```{r}
compute_regression_coefs <- function(x, y, z, data) {
  # recupera i nomi delle strutture
  tracts <- unique(data$structure)
  
  tract_names <- c()
  
  # not adjusted
  age_coefs_na <- c()
  age_low_cis_na <- c()
  age_upper_cis_na <- c()
  age_t_na <- c()
  age_P_na <- c()
  
  # adjusted age coeff
  age_coefs_adj <- c()
  age_low_cis_adj <- c()
  age_upper_cis_adj <- c()
  age_t_adj <- c()
  age_P_adj <- c()
    
  # adjusted covariate coeff
  cov_coefs_adj <- c()
  cov_low_cis_adj <- c()
  cov_upper_cis_adj <- c()
  cov_t_adj <- c()
  cov_P_adj <- c()
  
  for (i in 1:length(tracts)) {
    tract_name <- tracts[i]
    tract <- data[structure == tract_name]

    formula_na = as.formula(paste(y, paste(c(x), collapse='+'), sep=' ~'))
    formula_adj = as.formula(paste(y, paste(c(x, z), collapse='+'), sep=' ~'))
    ctrl <- lmrob.control(refine.tol=1e-5)
    m_na <- lmrob(formula_na, data=tract, control=ctrl)
    m_adj <- lmrob(formula_adj, data=tract, control=ctrl)
    
    tract_names <- c(tract_names, tract_name)
    
    # extract coefficients
    m_na_coeffs <- summary(m_na) %>% coef()
    m_adj_coeffs <- summary(m_adj) %>% coef()
    
    # extract confidence intervals
    m_na_cis <- confint(m_na)
    m_adj_cis <- confint(m_adj)

    # non adjusted age coefficient
    age_coefs_na <- c(age_coefs_na, round(m_na_coeffs["age", "Estimate"], 6))
    age_low_cis_na <- c(age_low_cis_na, round(m_na_cis["age", "2.5 %"], 5))
    age_upper_cis_na <- c(age_upper_cis_na, round(m_na_cis["age", "97.5 %"], 5))
    age_t_na <- c(age_t_na, round(m_na_coeffs["age", "t value"], 3))
    age_P_na <- c(age_P_na, round(m_na_coeffs["age", "Pr(>|t|)"], 4))
    
    # adjusted age coefficient
    age_coefs_adj <- c(age_coefs_adj, round(m_adj_coeffs["age", "Estimate"], 5))
    age_low_cis_adj <- c(age_low_cis_adj, round(m_adj_cis["age", "2.5 %"], 5))
    age_upper_cis_adj <- c(age_upper_cis_adj, round(m_adj_cis["age", "97.5 %"], 5)) 
    age_t_adj <- c(age_t_adj, round(m_adj_coeffs["age", "t value"], 2))
    age_P_adj <- c(age_P_adj, round(m_adj_coeffs["age", "Pr(>|t|)"], 4))
      
    # adjusted RD coefficient
    cov_coefs_adj <- c(cov_coefs_adj, round(m_adj_coeffs[z, "Estimate"], 5))
    cov_low_cis_adj <- c(cov_low_cis_adj, round(m_adj_cis[z, "2.5 %"], 5))
    cov_upper_cis_adj <- c(cov_upper_cis_adj, round(m_adj_cis[z, "97.5 %"], 5))  
    cov_t_adj <- c(cov_t_adj, round(m_adj_coeffs[z, "t value"], 2))
    cov_P_adj <- c(cov_P_adj, round(m_adj_coeffs[z, "Pr(>|t|)"], 4))  
  }
  
  regre_coeffs <- data.frame(tract_names, age_coefs_na, age_low_cis_na, 
                             age_upper_cis_na, age_t_na, age_P_na, age_coefs_adj, 
                             age_low_cis_adj, age_upper_cis_adj, age_t_adj, 
                             age_P_adj, cov_coefs_adj, cov_low_cis_adj, 
                             cov_upper_cis_adj, cov_t_adj, cov_P_adj)  
  
  filename <- sprintf("regression_coefs_robustlm_%s_age_%s.xlsx", y, z)
  write.xlsx(regre_coeffs, filename)
}
```

## RD

### Age and volume
```{r}
compute_regression_coefs(x='age', y='rd_mean', z='volume', data=df)
```

### Age and avg length
```{r}
compute_regression_coefs(x='age', y='rd_mean', z='avg_streamline_length', data=df)
```

### Age and density
```{r}
compute_regression_coefs(x='age', y='rd_mean', z='vol_length_ratio', data=df)
```

## AD

### Age and volume
```{r}
compute_regression_coefs(x='age', y='ad_mean', z='volume', data=df)
```

### Age and avg length
```{r}
compute_regression_coefs(x='age', y='ad_mean', z='avg_streamline_length', data=df)
```

### Age and density
```{r}
compute_regression_coefs(x='age', y='ad_mean', z='vol_length_ratio', data=df)
```

## MD
### Age and volume
```{r}
compute_regression_coefs(x='age', y='md_mean', z='volume', data=df)
```

### Age and avg length
```{r}
compute_regression_coefs(x='age', y='md_mean', z='avg_streamline_length', data=df)
```

### Age and density
```{r}
compute_regression_coefs(x='age', y='md_mean', z='vol_length_ratio', data=df)
```




# Mosaic plots of morphologic measures vs Age
## Length ~ Age
```{r}
ggplot(df, aes(x=age, y=avg_streamline_length)) +
    stat_smooth(method=function(formula, data, weights) lmrob(formula,data)) +
    stat_cor(method='spearman',
             p.accuracy=0.001, 
             r.accuracy=0.01,
             size=10) +
    facet_wrap(~structure, ncol=6, scales='free') +
    theme(strip.text = element_text(size=20))
```

## Density ~ Age
```{r}
ggplot(df, aes(x=age, y=vol_length_ratio)) +
    stat_smooth(method=function(formula, data, weights) lmrob(formula,data)) +
    stat_cor(method='spearman',
             p.accuracy=0.001, 
             r.accuracy=0.01,
             size=10) +
    facet_wrap(~structure, ncol=6, scales='free') +
    theme(strip.text = element_text(size=20))
```

## Volume ~ Age
```{r}
ggplot(df, aes(x=age, y=volume)) +
    stat_smooth(method=function(formula, data, weights) lmrob(formula,data)) +
    stat_cor(method='spearman',
             p.accuracy=0.001, 
             r.accuracy=0.01,
             size=10) +
    facet_wrap(~structure, ncol=6, scales='free') +
    theme(strip.text = element_text(size=20))
```


# New version of regression's mosaic plot with covariates: 
MD, AD, RD vs Age and correction for Volume, Density and Length


```{r}
scaleFUN <- function(x) sprintf("%.3f", x)

plot_rob_reg_with_morph <- function(model, models_adj, cov_labels, xvar, xlabel, ylabel, title) {
  # Predictions for non-adjusted model
  preds_na <- ggpredict(model, xvar)
  
  # Start building the predictions dataframe
  # 1. Add predictions non non adjusted model
  preds <- data.frame(preds_na$x, preds_na$predicted, preds_na$predicted)
  colnames(preds) <- c('x', 'preds_na', 'preds_adj')
  preds[['Legend']] <- 'No adj.'
  
  # for each adjusted model, extract predictions and add to preds dataframe
  for (i in 1:length(models_adj)) {
    cova <- cov_labels[i]
    model_adj <- models_adj[[i]]
    
    # extract and add predictions for current adjusted model
    preds_adj <- ggpredict(model_adj, xvar)  
    preds_adj <- data.frame(preds_na$x, preds_na$predicted, preds_adj$predicted)
    colnames(preds_adj) <- c('x', 'preds_na', 'preds_adj')
    preds_adj[['Legend']] <- cova
    
    preds <- rbind(preds, preds_adj)    
  }
  
  # convert Legend column to factor
  preds$Legend <- as.factor(preds$Legend)
  
  # reorder levels of Legend
  levels(preds$Legend) <- c('Volume', 'Density', 'Length', 'No Adj.')
  
  min_y <- min(preds$preds_na, preds$preds_adj)
  max_y <- max(preds$preds_na, preds$preds_adj)
  plt <- ggplot(preds, aes(x=x, y=preds_adj, colour=Legend, group=Legend)) + 
    scale_y_continuous(labels=scaleFUN, 
                       breaks=round(seq(min_y, max_y, by = (max_y - min_y)/4),3)) +
    geom_line() +
    labs(x = xlabel, y = ylabel, title=title) +
    theme_classic() + 
    theme(legend.position="bottom")
  
  return(plt)
}
```

## Make MD plots
```{r}
covariates <- c('volume', 'vol_length_ratio', 'avg_streamline_length')
cov_labels <- c('Volume', 'Density', 'Length')

y <- 'md_mean'
x <- 'age'
formula_na = as.formula(paste(y, paste(c(x), collapse='+'), sep=' ~'))
ctrl <- lmrob.control(refine.tol=1e-5)

md_plots <- list()

for (t in 1:length(tracts)) {
  tract_name <- tracts[t]
  tract <- df[structure == tract_name]
  
  m_na <- lmrob(formula_na, data=tract, control=ctrl)
  
  models_adj <- list()
  
  for (i in 1:length(covariates)) {
    cova <- covariates[i]
    formula_adj = as.formula(paste(y, paste(c(x, cova), collapse='+'), sep=' ~'))
    m_adj <- lmrob(formula_adj, data=tract, control=ctrl)
    summary(m_adj)
    
    models_adj[[i]] <- m_adj
  }
  
  plt <- plot_rob_reg_with_morph(model=m_na, 
                          models_adj=models_adj, 
                          cov_labels=cov_labels, 
                          xvar='age',
                          xlabel='Age (years)',
                          ylabel=bquote('MD ' (mm/s^2)),
                          title=tract_name
                          )  
  md_plots[[t]] <- plt
}
```

## Make AD plots
```{r}
covariates <- c('volume', 'vol_length_ratio', 'avg_streamline_length')
cov_labels <- c('Volume', 'Density', 'Length')

y <- 'ad_mean'
x <- 'age'
formula_na = as.formula(paste(y, paste(c(x), collapse='+'), sep=' ~'))
ctrl <- lmrob.control(refine.tol=1e-5)

ad_plots <- list()

for (t in 1:length(tracts)) {
  tract_name <- tracts[t]
  tract <- df[structure == tract_name]
  
  m_na <- lmrob(formula_na, data=tract, control=ctrl)
  
  models_adj <- list()
  
  for (i in 1:length(covariates)) {
    cova <- covariates[i]
    formula_adj = as.formula(paste(y, paste(c(x, cova), collapse='+'), sep=' ~'))
    m_adj <- lmrob(formula_adj, data=tract, control=ctrl)
    summary(m_adj)
    
    models_adj[[i]] <- m_adj
  }
  
  plt <- plot_rob_reg_with_morph(model=m_na, 
                          models_adj=models_adj, 
                          cov_labels=cov_labels, 
                          xvar='age',
                          xlabel='Age (years)',
                          ylabel=bquote('AD ' (mm/s^2)),
                          title=tract_name
                          )  
  ad_plots[[t]] <- plt
}
```

## Make RD plots
```{r}
covariates <- c('volume', 'vol_length_ratio', 'avg_streamline_length')
cov_labels <- c('Volume', 'Density', 'Length')

y <- 'rd_mean'
x <- 'age'
formula_na = as.formula(paste(y, paste(c(x), collapse='+'), sep=' ~'))
ctrl <- lmrob.control(refine.tol=1e-5)

rd_plots <- list()

for (t in 1:length(tracts)) {
  tract_name <- tracts[t]
  tract <- df[structure == tract_name]
  
  m_na <- lmrob(formula_na, data=tract, control=ctrl)
  
  models_adj <- list()
  
  for (i in 1:length(covariates)) {
    cova <- covariates[i]
    formula_adj = as.formula(paste(y, paste(c(x, cova), collapse='+'), sep=' ~'))
    m_adj <- lmrob(formula_adj, data=tract, control=ctrl)
    summary(m_adj)
    
    models_adj[[i]] <- m_adj
  }
  
  plt <- plot_rob_reg_with_morph(model=m_na, 
                          models_adj=models_adj, 
                          cov_labels=cov_labels, 
                          xvar='age',
                          xlabel='Age (years)',
                          ylabel=bquote('RD ' (mm/s^2)),
                          title=tract_name
                          )  
  rd_plots[[t]] <- plt
}
```

```{r}
plots <- cbind(as.vector(md_plots), 
                as.vector(ad_plots), 
                as.vector(rd_plots))
```


```{r}
yaxis_title_size <- 8
xaxis_title_size <- 8
plots_step <- 8

theme_fst_col <- theme(axis.text.y = element_text(size=5),
                       axis.ticks.y = element_blank(),
                       axis.title.y = element_blank(),
                       axis.text.x = element_blank(),
                       axis.ticks.x = element_blank(),
                       axis.title.x = element_blank(),
                       plot.margin = margin(r = 1) )

bottom_left <- theme(axis.text.y = element_text(size=5),
                     axis.ticks.y = element_blank(),
                     axis.title.y = element_blank(),
                     axis.text.x = element_text(size=5),
                     axis.ticks.x = element_blank(),
                     axis.title.x = element_text(size=xaxis_title_size),
                     plot.margin = margin(r = 1) )


inner_theme <- theme(axis.text.y = element_text(size=5),
                     axis.ticks.y = element_blank(),
                     axis.title.y = element_blank(),
                     axis.text.x = element_blank(),
                     axis.ticks.x = element_blank(),
                     axis.title.x = element_blank(),
                     plot.margin = margin(r = 1, l = 1) )

bottom_inner_theme <- theme(axis.text.y = element_text(size=5),
                     axis.ticks.y = element_blank(),
                     axis.title.y = element_blank(),
                     axis.text.x = element_text(size=5),
                     axis.ticks.x = element_blank(),
                     axis.title.x = element_text(size=xaxis_title_size),
                     plot.margin = margin(r = 1, l = 1) )


for (i in seq(1, nrow(plots)-1, 1)) {
  # apply theme_fst_col to first plot
  plots[i,][[1]] <- plots[i,][[1]] + theme_fst_col
  
  # apply inner_theme to second and third plots
  plots[i,][[2]] <- plots[i,][[2]] + inner_theme
  plots[i,][[3]] <- plots[i,][[3]] + inner_theme
}

# treat last row of plots of each of part the same
for (i in seq(from=plots_step, to=nrow(plots), by=plots_step)) {
  plots[i,][[1]] <- plots[i,][[1]] + bottom_left

  # apply bottom_inner_theme to last two plots
  plots[i,][[2]] <- plots[i,][[2]] + bottom_inner_theme
  plots[i,][[3]] <- plots[i,][[3]] + bottom_inner_theme
}


for (i in 1:length(plots)) {
  plots[[i]] <- plots[[i]] + theme(plot.title = element_text(size=8),
                                  plot.margin = unit(c(0.1,0.15,0.1,0.15), "cm"))
}

# reduce each row of plots adding each with +
reductions <- c()
for (r in 1:nrow(plots)) {
  reductions[[r]] <- reduce(plots[r,], `+`)
}


# stack triplets by reducing each with /
for (i in seq(from=1, to=length(reductions), by=8)) {
  to <- (i+8)-1
  part <- (reduce(reductions[i:to], `/`) + plot_layout(guides="collect")) & theme(legend.position = "bottom")
  ggsave(sprintf("%s.pdf",i),  # jpg, png, eps, tex, etc.
       plot = part, # or an explicit ggplot object name,
       width = 4, height = 8, 
       units = "in", # other options c("in", "cm", "mm"), 
       dpi = 300)
}
```

